|

### 2021-08-03 03-Chp-Systematic-Review.Rmd
End of day:

Without our framework and as part of the risk of bias assessments reported in \@ref(risk-of-bias-subheading), Itest test

>>>>

Without our framework and as part of the risk of bias assessments reported in \@ref(risk-of-bias-subheading), I attempted to records the direction of the bias, so that it could feed into the triangulation.

### 2021-08-03 04-Chp-CPRD-Analysis.Rmd
End of day:

@@

This section will expand on a potential explanation for the observed results detailed above. However, as the comparison of evidence across different sources is the aim of the triangulation exercise presented in later chapters, the section will not provide a detailed comparison with other published literature, except where needed to illustrate a methodological point. For a comparison of the result presented above with those from the systematic review (Chapter \@ref(hold)) and the individual patient data analysis (Chapter \@ref(hold)), see Section \@ref(hold).

>>>>

<!-- IDEA Could frame this as providing evidence that the effect is not due to statins acting on lipid levels, as then would not expect difference (unless all very hard acting statins are in the lipophilic type) -->

@@

This section will expand on a potential explanation for the observed results detailed above. However, as the comparison of evidence across different sources is the aim of the triangulation exercise presented in later chapters, the section will not provide a detailed comparison with other published literature, except where needed to illustrate a methodological point. For a comparison of the result presented above with those from the systematic review (Chapter \@ref(sys-rev-heading)) and the individual patient data analysis (Chapter \@ref(ipd-heading)), see Section \@ref(discussion-heading).

### 2021-08-03 manuscripts/cprd/manuscript.Rmd
End of day:

* Main text word count (max 3000): 2999
* Abstract word count (max 250): 244

@@

__Background:__ There is some evidence that circulating blood lipids play a role in the development of Alzheimer's disease (AD) and dementia. As a result, these modifiable risk factors could be targeted by existing lipid-regulating agents, including statins, for the prevention of dementia. Here, we test the association between lipid-regulating agents and subsequent risk of dementia and related conditions in the Clinical Practice Research Datalink (CPRD), a large-scale electronic health record database.

__Methods:__ A cohort study design using data from the CPRD, routinely collected between January 1995 and March 2016, was performed. Cox proportional hazard models, allowing for a time-varying treatment indicator, were used to estimate the association between seven lipid-regulating drug classes (vs. no drug) and five dementia outcomes (all-cause, vascular and other dementia, and probable and possible Alzheimer's disease).

__Results:__ We analyzed `r totaltext` with a total follow-up of `r comma(total_followup)` participant-years (median: `r fu_text`). We found little evidence that lipid-regulating agents were associated with risk of Alzheimer's disease (probable `r probad_text`; possible `r possad_text`), but there was evidence of an increased risk of all-cause (`r anydem_text`), vascular (`r vasdem_text`) and other dementia (`r othdem_text`). Evidence from a number of control outcomes (ischaemic heart disease `r ihd_text`; backpain `r backpain_text` and diabetes `r dm_type2_text`) indicated the presence of substantial residual confounding related to vascular factors.

__Conclusion:__ We found little evidence that lipid-regulating agents were associated with reduced on Alzheimer’s disease risk. There was some evidence of an increased the risk of all-cause, vascular and other dementia, likely the result of residual confounding by indication.

@@

The use of electronic health data for epidemiological research has several advantages.[@casey2016] As the data are collected through the routine care of a large cohort, they allow for nested cohort studies using sample sizes and time-scales which would be infeasible using traditional methods. In addition, data are collected for care provision and without a specific research question in mind,  providing a holistic picture of a patient and their health experience. This provides vital data on a range of potential confounders which can be incorporated into an analysis.

@@

The index date for a participant was defined as the date where the first relevant code or test result (as detailed above) was recorded on their clinical record. Participants were followed up until the earliest of: an outcome of interest; death; end of follow-up (29 February 2016); last registration date with their GP practice; last CPRD collection date for their practice. Participants were removed from our sample if they were less than 40 years of age, had less than 12 months of "research quality" data prior to their index date, were simultaneously prescribed more than one LRA (due to the difficulty of assigning these patients to a single exposure group), or were diagnosed with dementia before or on the date of the index event.

@@

A total of `r format(attrition[14,1], big.mark = ",")` participants met the inclusion criteria for our cohort (See Supplementary Figure 2 for the attrition flowchart), with a total follow-up of `r comma(total_followup)` patient years at risk. Most participants were included in the cohort due to an elevated test result (`r index_event_text`). The median age at index was `r age_text` and participants were followed up for a median of `r fu_text`. During follow-up, an all-cause dementia diagnosis was recorded for `r format(p1[1,2], big.mark = ",")` patients  (`r format(p1[2,2], big.mark = ",")` probable Alzheimer's disease, `r format(p1[3,2], big.mark = ",")` possible Alzheimer's disease, `r format(p1[4,2], big.mark = ",")` vascular dementia, `r format(p1[5,2], big.mark = ",")` other dementia). The distribution of baseline characteristics across the drug classes can be seen in Table \@ref(tab:cprdCharacteristics-table).

@@

For our control outcomes (Supplementary Figure 6), there was some evidence that patients prescribed a lipid-regulating agent had increased risks of back pain (`r backpain_text`), ischaemic heart disease (`r ihd_text`) and Type 2 diabetes (`r dm_type2_text`).

@@

A recent Cochrane Review identified two randomized trials comparing treatment with statins versus non-treatment for the prevention of dementia, only one of which presented information on the incidence of dementia.[@mcguinness2016] This study (Heart Protection Study) showed no effect of treatment with simvastatin on all-cause dementia risk (OR: 1.00,  95%CI:0.61-1.65),[@heartprotectionstudycollaborativegroup2002] but concerns were raised over the diagnostic criteria used. A meta-analysis of 30 observational studies found a reduced risk of all-cause dementia was associated with statin treatment (RR 0.83, 95%CI: 0.79–0.87).[@poly2020]

These sources of evidence conflict with the findings of our analysis, where statin use was associated with an increased risk of all-cause dementia (`r anydem_text`). However, some of the included studies in the meta-analysis specifically exclude vascular dementia from the definition of all-cause dementia,[@chao2015] which may lead to an artificial protective effect of statins on all-cause dementia and limit the ability for comparison between studies.

@@

@@

__Statins and non-Alzheimer's disease dementia__

@@

Our study is likely to be subject to confounding by indication, which could provide a potential explanation for the observed increased risk of vascular and other dementia with lipid regulating agent use. Supporting evidence for this interpretation comes from a variety of sources, including the results of the control outcome analyses. The slight harmful effect for the backpain outcome is substantially smaller than that observed for the ischaemic heart disease outcome, indicating that the majority of the uncontrolled confounding is likely related to vascular factors. Additionally, we obtained the expected harmful result for Type 2 diabetes, where statins' mechanism of action on this outcome is unlikely to be vascular.[@collins2016a; @macedo2014] Further supporting evidence comes from the increasingly harmful effect when moving from the Probable/Possible Alzheimer's disease to other dementia to vascular dementia outcomes, indicating that confounding by indication likely increases as the proportion of cases with a vascular component increases. A review of other available literature suggests that this observation (a harmful effect of lipid regulating agents on vascular-related outcome due to confounding by indication) is not unusual. Using a conventional epidemiological technique, a previous analysis also found an increased risk of coronary heart disease (analogous to the ischaemic heart disease outcome used in our analysis) in those taking statins (`r estimate(1.31,1.04, 1.66, type = "HR")`).[@danaei2013b]

@@

We have attempted to account for important sources of bias in our analysis and provide a comparison with other available literature. However, there is a strong potential for unmeasured confounding, misclassification and reverse causation, which may relate to the unexpected increase in risk of vascular dementia associated with statin use. Future research should aim to address these potential biases and, while it may be costly in terms of time and resources, a large scale, long-term randomized controlled trial would provide useful additional information on the effect of lipid-regulating agents on the risk of dementia and related outcomes.

@@

@@

_Figure 1: Hazard ratios produced by the primary analyses of CPRD data, stratified by dementia outcome and drug class. All results were obtained using the fully adjusted model and participant age as the time scale._

>>>>

* Main text word count (max 4000): 3104
* Abstract word count (max 250): 250

@@

__Check TODOs!__

<!-- TODO Fix total length of follow-up between text and table 2 -->

__Background:__ There is some evidence that circulating blood lipids play a role in the development of Alzheimer's disease (AD) and dementia. As a result, these modifiable risk factors could be targeted by existing lipid-regulating agents, including statins, for the prevention of dementia. Here, we test the association between lipid-regulating agents and risk of dementia and related conditions in the Clinical Practice Research Datalink (CPRD), an electronic health record database.

__Methods:__ A cohort study using data from the CPRD, routinely collected between January 1995 and March 2016, was performed. Multivariable Cox proportional hazard models, allowing for a time-varying treatment indicator, were used to estimate the association between seven lipid-regulating drug classes (vs. no drug) and five dementia outcomes (all-cause, vascular and other dementia, and probable and possible Alzheimer's disease).

__Results:__ We analyzed `r totaltext` with a total follow-up of `r comma(total_followup)` participant-years (median: `r fu_text`). We found little evidence that lipid-regulating agents were associated with Alzheimer's disease (probable `r probad_text`; possible `r possad_text`), but there was evidence of an increased risk of all-cause (`r anydem_text`), vascular (`r vasdem_text`) and other dementia (`r othdem_text`). Evidence from a number of control outcomes (ischaemic heart disease `r ihd_text`; backpain `r backpain_text`; and Type 2 diabetes `r dm_type2_text`) indicated the presence of substantial residual confounding by indication.

__Conclusion:__ We found little evidence that lipid-regulating agents were associated with reduced Alzheimer’s disease risk. There was some evidence of an increased the risk of all-cause, vascular and other dementia, likely due to residual confounding by indication.

@@

The use of electronic health data for epidemiological research has several advantages.[@casey2016] As the data are collected through the routine care of a large cohort, they allow for nested cohort studies using sample sizes and time-scales which would be infeasible using traditional methods. In addition, data are collected for care provision and without a specific research question in mind,  providing a holistic picture of a patient and their health experience. While they may not contain information on all potentially important confounders,[@farmer2018] electronic health records provide routine data on common covariates for millions of participants.

@@

The index date for a participant was defined as the date where the first relevant code or test result (as detailed above) was recorded on their clinical record. Participants were followed up until the earliest of: an outcome of interest; death; end of follow-up (29 February 2016); last registration date with their GP practice; or the last CPRD collection date for their practice. Participants were removed from our sample if they were less than 40 years of age, had less than 12 months of "research quality" data prior to their index date, were simultaneously prescribed more than one LRA (due to the difficulty of assigning these patients to a single exposure group), or were diagnosed with dementia before or on the date of the index event.

@@

A total of `r format(attrition[14,1], big.mark = ",")` participants met the inclusion criteria for our cohort (see Supplementary Figure 2 for the attrition flowchart), with a total follow-up of `r comma(total_followup)` patient years at risk. Most participants were included in the cohort due to an elevated test result (`r index_event_text`). The median age at index was `r age_text` and participants were followed up for a median of `r fu_text`. During follow-up, an all-cause dementia diagnosis was recorded for `r format(p1[1,2], big.mark = ",")` patients  (`r format(p1[2,2], big.mark = ",")` probable Alzheimer's disease, `r format(p1[3,2], big.mark = ",")` possible Alzheimer's disease, `r format(p1[4,2], big.mark = ",")` vascular dementia, `r format(p1[5,2], big.mark = ",")` other dementia). The distribution of baseline characteristics across the drug classes can be seen in Table \@ref(tab:cprdCharacteristics-table).

@@

For our control outcomes (Supplementary Figure 6), there was some evidence that patients prescribed a lipid-regulating agent had an increased risk of back pain (`r backpain_text`), ischaemic heart disease (`r ihd_text`) and Type 2 diabetes (`r dm_type2_text`).

@@

A recent Cochrane Review identified two randomized trials comparing treatment with statins versus non-treatment for the prevention of dementia, only one of which presented information on the incidence of dementia.[@mcguinness2016] This study (Heart Protection Study) showed no effect of treatment with simvastatin on all-cause dementia risk (OR: 1.00,  95%CI:0.61-1.65),[@heartprotectionstudycollaborativegroup2002] but concerns were raised over the diagnostic criteria used. A meta-analysis of 30 observational studies found a reduced risk of all-cause dementia was associated with statin treatment (RR 0.83, 95%CI: 0.79–0.87).[@poly2020] This conflicts with the findings of our analysis, where statin use was associated with an increased risk of all-cause dementia (`r anydem_text`). However, some of the included studies in the meta-analysis specifically exclude vascular dementia from the definition of all-cause dementia,[@chao2015] which may limit the ability for comparison with the results of this analysis.

Additionally, a previous analysis of the THIN EHR database using a propensity score matching analysis found a protective effect of statins on all-cause dementia (HR:0.81, 95%CI:0.69-0.96),[@smeeth2009]. Differences in dementia coding over time, in addition to a different set of covariates adjusted for in the analysis, may go someway to explaining the discrepancy.

@@

<!-- TODO Need to cite Smeeth paper and discuss -->

@@

__Statins and vascular/other dementia__

@@

Our study is likely to be subject to confounding by indication, which could provide a potential explanation for the observed increased risk of vascular and other dementia with lipid regulating agent use. Supporting evidence for this interpretation comes from a variety of sources, including the results of the control outcome analyses. The slight harmful effect for the backpain outcome is substantially smaller than that observed for the ischaemic heart disease outcome, indicating that the majority of the uncontrolled confounding is likely related to vascular factors. Additionally, we obtained the expected harmful result for Type 2 diabetes, where statins' mechanism of action on this outcome is unlikely to be vascular.[@collins2016a; @macedo2014] Further supporting evidence comes from the increasingly harmful effect when moving from the Probable/Possible Alzheimer's disease to other dementia to vascular dementia outcomes, indicating that confounding by indication likely increases as the proportion of cases with a vascular component increases. A review of other available literature suggests that this observation (a harmful effect of lipid regulating agents on vascular-related outcome due to confounding by indication) is not unusual. Using a conventional epidemiological technique, a previous analysis also found an increased risk of coronary heart disease (analogous to the ischaemic heart disease outcome used in our analysis) in those taking statins (`r estimate(1.31,1.04, 1.66, type = "HR")`).[@danaei2013b] Following control for confounding by indication in that study through use of a trial emulation analysis, the expected protective effect of statins was observed.

@@

We have attempted to account for important sources of bias in our analysis and provide a comparison with other available literature. However, there is a strong potential for unmeasured confounding, misclassification and reverse causation, which may relate to the unexpected increase in risk of vascular dementia associated with statin use. Other analytical methods. Future research should aim to address these biases, potentially via a trial emulation analysis.

@@

<!----------------------------------------------------------------------------->
(ref:followUp-caption) Summary of person-time at risk and number of events by exposure group.

```{r followUp-table, message=FALSE, results="asis", echo = FALSE}
```
<!----------------------------------------------------------------------------->

@@

_Figure 1: Results from the primary analyses of CPRD data comparing prescription of an lipid-regulating drug with no prescription, stratified by dementia outcome and drug class. All results were obtained using the fully adjusted model and participant age as the time scale._
